---
title: "D Plunkett 607 Week 1"
author: "Diana Plunkett"
date: "1/30/2022"
output: html_document
---
<hr/>  
## Overview

The data set used here is from the article "The (Very) Long Tail of Hurricane Recovery".    
https://projects.fivethirtyeight.com/sandy-311/    

The article discussed how nearly 5 years after Sandy, 311 was still getting calls from New Yorkers asking for information.  

The data is pulled from NYC Open Data, and is structured with a row for each day from Oct 29, 2012 though Sept 18, 2017.  The columns consist of a date, followed by a columns for each NYC Agency to which calls were referred. 

Although the stream chart in the article was beautiful, and perfect for illustrating the long tail, I had a hard time comparing which agencies got the most calls each year.  Below I restructure and chart the data to do that.  

### Bring in libraries and data

```{r}
library(tidyverse)
library(magrittr)

sandy_data <-read_csv("https://raw.githubusercontent.com/dianaplunkett/607/main/sandy-311-calls-by-day.csv", show_col_types = FALSE)
```

### Pivot the data
Pivot the data to get the NYC Agencies to be the data in a column, rather than the name of a column. 

```{r}
sandy_data %<>% pivot_longer(!date, names_to = "agency", values_to = "calls_count")

```

### Summerize and tidy the data     
* pull year from date so we can then ...    
* summarize by year and agency    
* filter "total" from agency, which was an original column, and is not an agency    
* also filter any row with calls= 0 (eg row where an agency had no calls that year) 

```{r}
# year is in the last 2 characters of the data field
sandy_data %<>% mutate(year = str_sub(sandy_data$date, -2,-1))

annual_sandy_data <- sandy_data %>% 
    group_by(year, agency) %>% 
    summarize(calls = sum(calls_count, na.rm = TRUE))

annual_sandy_data %<>% filter(agency !="total")
annual_sandy_data %<>% filter(calls !=0 )
```
### Augment the data    
* calculate the total per year    
* and then get the % of total for each agency for each year    

```{r}
year_total =  annual_sandy_data %>% 
    group_by(year) %>% 
    summarize(year_calls = sum(calls, na.rm = TRUE))

annual_sandy_data <- merge(x=annual_sandy_data, y=year_total, by='year')
annual_sandy_data %<>% mutate (percent_of_total = calls / year_calls)


```
### Keeping only the top 5 rows per year

The function here gets the agencies with the top 5 largest 'percent of total's each year.  I am sure there is a clever way to call the function for each year and union the resulting data, and I tried to do that with a for loop, but my R skills are not yet to that level.  I left in a bit of what I attempted as a comment so when I learn more about for loops, I can figure it out. 

The code below works, it is just not very elegant.

```{r}
get_top5 <- function(x) {
    top5 <-annual_sandy_data %>% 
    filter(year == x) %>%
    group_by(year)%>%
    arrange(desc(percent_of_total)) %>%
    head(5)
    return(top5)
}

#tried to use a for loop to union the top 5 for each year, but could not quite get it
# tried variations on the below
#for (i in 12:17) {
#   print(i)
#    union (just_top5, get_top5(i))
#}
#so I "cheated" and just did it manually. Works for a data set of this size.
#and since loops were not the focus of this work, assuming this is OK.

just_top5_12 <- get_top5(12)
just_top5_13 <- get_top5(13)
just_top5_14 <- get_top5(14)
just_top5_15 <- get_top5(15)
just_top5_16 <- get_top5(16)
just_top5_17 <- get_top5(17)
U12_13 <-union(just_top5_13, just_top5_12)
U14_15 <-union(just_top5_14, just_top5_15)
U16_17 <-union(just_top5_16, just_top5_17)
U12thru15<-union(U12_13, U14_15)
all_top5 <- union(U12thru15, U16_17)
```

### Plot it

Now that we have arranged the data, let's see what we can learn about which agencies were involved with the 311 calls each year (limiting it to only the top 5 agencies per year.    

```{r}
color_map<-tibble(type= c('12', '13', '14','15', '16', '17'), 
                  color = c('#154360', '#2980b9', '#5aa8dc', 
                            '#a9cce3', '#aed6f1',  '#ebf5fb' ))

all_top5 %>% ggplot() +
    geom_col(mapping = aes(x=agency, y=percent_of_total, fill=year)) + 
    scale_fill_manual(values=color_map$color) +
    coord_flip() +
    ggtitle("Top 5 Agencies involved in 311 calls about Huricane Sandy by Year") + 
    ylab('% total calls made in a year')+
    xlab("Agency")
```

### Conclusion    

Pivoting the data to make the agency the focus, summarizing by year/agency and then calculating the % of a year's total for each agency let us see which agencies got the most calls each year.  

By limiting the data to the top 5 agencies each year, and then plotting this, one can see:   
* NYS Emergency Management (NYSEMERGENCYMG) got almost 1/4 of the calls in the 'tail' years of 2015, 2016 and 2017    
* The Dept of Finance (DOF), Dept of Buildings (DOB) and City Hall (CHall), along with NYS Emergency Management got calls in most years.    

*Note: I considered "translating" the Agency abbreviation, but given some the the long names, it would clutter the chart.  Therefore I decided to translate here in the conclusions.*  


