---
title: "D Plunkett 607 Project 1"
author: "Diana Plunkett"
date: "2/21/2022"
output: html_document
---
## Overview




### Load libraries and other set up

```{r Libraries, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(magrittr)
```

### Load file

```{r, Load-File, message=FALSE, warning=FALSE}
Chess<- read_delim(
            "https://raw.githubusercontent.com/dianaplunkett/607/main/tournamentinfo.txt",
           delim = "|",
           col_names = FALSE,
           trim_ws = TRUE, 
           skip = 4, 
           show_col_types = FALSE)
```


### Initial file cleanup
- remove the final empty column 11
- get rid of every third row with all dashes in column X1 and NA in other columns

```{r Init-clean, message=FALSE, warning=FALSE}
Chess %<>% 
    select( -X11) %>%
    drop_na (X2)
```

### Now split into Odd and Even rows & bind_col
The data now has "row pairs" with a single observation in 2 rows (eg Player 1 results are on row 1 and 2, Player 2 results are on row 3 and 4 etc).    

Split the data into Odd and Even rows, then use bind_col to put each pair of rows back into a single row.  

```{r odd-even, message=FALSE, warning=FALSE}
row_odd <- seq_len(nrow(Chess)) %% 2
Odd <- Chess[row_odd ==1,]
Even <- Chess[row_odd ==0,]
Chess2 <- bind_cols(Odd, Even)
```

### Filter to needed columns, rename, and separate and mutate    
- remove the unneeded columns that come in from the second row of the "pairs"    
- Now that we have just what we need, rename the remaining columns to be more meaningful    
- separate the column with the playerID/pre/post ratings.  We only need the prerating here, so the separate focuses on isolating those 4 characters in "pre".  "id" and "post" end up with extraneous characters, but does not matter as we do not use them.    
- in fact, we will remove those 2 unneeded columns, "id" and "post"    
- separate the rd1-7 columns into cd1-cd7 and opp1-opp7 - separating the W/L/D/H/B codes from the opponent # 
- then make 2 new new columns:  cds=a list of cd1-cd7 and opps= a list of opp1-opp7

At the end of this step, we have just the data we need, and everything about a particular player is in a single row.  

```{r final-clean, message=FALSE, warning=FALSE}
Chess2 %<>%
    select (-X3...13, -X4...14, -X5...15, -X6...16, -X7...17, -X8...18, -X9...19, -X10...20) 

colnames(Chess2) <- c('player', 'name', 'total_pts', 'rd1', 'rd2', 'rd3', 'rd4', 'rd5', 'rd6','rd7', 'state', 'idprepost')

Chess2 <- separate(Chess2, col=idprepost, into = c("id", "pre", "post"), sep=c(14,18))

Chess2 %<>%
    select ( -id, -post) %>% 
    mutate(pre = as.integer(pre))

extract_opp <- function(rdx) {
    as.numeric(str_extract(rdx, "(\\d)+"))
}

Chess2$opp1 <-extract_opp(Chess2$rd1)
Chess2$opp2 <-extract_opp(Chess2$rd2)
Chess2$opp3 <-extract_opp(Chess2$rd3)
Chess2$opp4 <-extract_opp(Chess2$rd4)
Chess2$opp5 <-extract_opp(Chess2$rd5)
Chess2$opp6 <-extract_opp(Chess2$rd6)
Chess2$opp7 <-extract_opp(Chess2$rd7)

Chess2 %<>% 
    rowwise %>%
    mutate( opps= list(c(opp1,opp2,opp3,opp4,opp5,opp6,opp7)))  %>%
    select (name, state, total_pts, pre, opps)

```

### Get at Avg Rating of Opponents & write to csv
For each row in the data (each player):    
        - sum to total of the pre tournament rating of the opponents    
        - count the number of games    
        - calculate the Average Opponent Rating

Set the final columns needed

Write to csv

```{r calc-avg-opp-rtg, message=FALSE, warning=FALSE}

for (i in 1:nrow(Chess2) )  {
    Chess2$oppTotal[i] <- sum(Chess2$pre[unlist(Chess2$opps[i])], na.rm=TRUE)
    Chess2$numGames[i] <- sum(!is.na(unlist(Chess2$opps[i])))
    Chess2$AvgOppRtg[i] <-as.integer(Chess2$oppTotal[i]/Chess2$numGames[i])
}

Output<-Chess2 %>% 
    select( - oppTotal, -numGames, -opps)

write.csv(Output, "dmp-chess-output.csv", quote = FALSE, row.names=FALSE)
    
```






