---
title: "D Plunkett 607 Week 3"
author: "Diana Plunkett"
date: "2/13/2022"
output: html_document
---

```{r}
library(tidyverse)
```

## 1
Using the 173 majors listed in fivethirtyeight.com’s College Majors dataset [https://fivethirtyeight.com/features/the-economic-guide-to-picking-a-college-major/], provide code that identifies the majors that contain either "DATA" or "STATISTICS"

```{r}
majors <-read_csv("https://raw.githubusercontent.com/dianaplunkett/607/main/majors-list.csv",
                  show_col_types = FALSE)
```

```{r}
majors %>% 
filter(str_detect(Major, "DATA|STATISTICS")) 

```

## 2.1
Write code that transforms the data below:    
    [1] "bell pepper"  "bilberry"     "blackberry"   "blood orange"    
    [5] "blueberry"    "cantaloupe"   "chili pepper" "cloudberry"    
    [9] "elderberry"   "lime"         "lychee"       "mulberry"       
    [13] "olive"        "salal berry"    
Into a format like this:    
c("bell pepper", "bilberry", "blackberry", "blood orange", "blueberry", "cantaloupe", "chili pepper", "cloudberry", "elderberry", "lime", "lychee", "mulberry", "olive", "salal berry")   

### After Wednesday's class, I think this is what is expected for this question:
Take in a file with 4 lines of text that look like:    

    [1] "bell pepper"  "bilberry"     "blackberry"   "blood orange"    
    [5] "blueberry"    "cantaloupe"   "chili pepper" "cloudberry"    
    [9] "elderberry"   "lime"         "lychee"       "mulberry"       
    [13] "olive"        "salal berry"    
    

And create a string that looks like:     

c("bell pepper", "bilberry", "blackberry", "blood orange", "blueberry", "cantaloupe", "chili pepper", "cloudberry", "elderberry", "lime", "lychee", "mulberry", "olive", "salal berry") 

```{r}

inputTXT<-readLines("https://raw.githubusercontent.com/dianaplunkett/607/main/produce.txt", 
                   n=-1,
                   warn=FALSE)
writeLines(inputTXT)

produce<-c(str_extract_all(inputTXT, "(\"\\w*[:blank:]?\\w*\")+", simplify=TRUE))
produce<-as.data.frame(produce)
produce <- as.data.frame(produce[!apply(produce, 1, function(x) any(x=="")),])
names(produce) [1]<-'NameList'
produce <- produce[order(produce$NameList),]


ProduceString <-paste0(produce, collapse=",", sep="" )
FinalString <-paste0("c(", ProduceString, ")", sep="")
writeLines(FinalString)

```


#### 2.2 The below is my earlier atempt at this, when I understood the question slightly differently.

```{r}
produce <-c("bell pepper", "bilberry", "blackberry", "blood orange", "blueberry", "cantaloupe", "chili pepper", "cloudberry", "elderberry", "lime", "lychee", "mulberry", "olive", "salal berry")

produce


```

```{r}
cat("c(", paste0(shQuote(produce, type="cmd"), collapse=", "), ")", sep="")

```

## 3

Describe, in words, what these expressions will match:

    (.)\1\1    
The above will match any three characters in a row, like 'ddd'.  The dot . matches any character, it is grouped with parentheses (.), and the \1 is a backreference to the group, and the second \1 is another backreference to the same group.    
    
    "(.)(.)\\2\\1"    
The above will match any two characters, followed by the reverse of those two characters, like 'dppd'.  We start with 2 dots (any 2 characters), each grouped (with parens), then the backreference to the 2nd group, then and a backreference to the first group.  

    (..)\1    
The above will match any two characters repeated: "dpdp".  A group of 2 characters, followed by a backreference.  
    
    "(.).\\1.\\1"
The above will match a string of 5 characters where the first, third and fifth character are the same: "dxdyd". A grouped character, then a dot (any char), then a backreference to the first group, another dot (any char) and another reference to the group.    

    "(.)(.)(.).*\\3\\2\\1"    
The above will match three characters followed by a wildcard of 0 or more characters and then the first three characters again in reveres order.  So all of the following will match: "dmppmd", "dmp8pmd", dmpabcdefghpmd".  Three individually grouped characters, a wildcard character which can be of any length including 0, and then back references to the 3rd, 2nd and 1st groups (in that order).

## 4
Construct regular expressions to match words that:

    Start and end with the same character.    
    
    "^(.)((.*\\1$)|\\1?$)" - ^ indicates starts with, (.) a group with any character, followed by a wildcard group    that ends ($) in the first character OR it is only one character.  The bit up to the | will find all words that    start and end with the same character that are more than one character long.  The part after the | (or) says    also include it if there are as few characters as the overall pattern to match (eg lazy, start ^ and end $ are (.)).    
```{r}
str_subset(words, "^(.)((.*\\1$)|\\1?$)")
```

    
    Contain a repeated pair of letters (e.g. "church" contains "ch" repeated twice.)        
    
    "([A-Za-z][A-Za-z]).*\\1" - the first group in parens is a letter (upper of lower case) followed by another       letter (upper or lower case), then a wildcard of 0 or more characters, and then a backreference to the first       group of 2 letters.  

```{r}
str_subset(words, "([A-Za-z][A-Za-z]).*\\1")
```

    Contain one letter repeated in at least three places (e.g. "eleven" contains three "e"s.)    
    
    "([a-z]).*\\1.*\\1" - group containing a letter, followed by a wild card of 0 or more characters, then a         backreference to the group letter, another wild card and then the group letter again. 

```{r}
str_subset(words, "([a-z]).*\\1.*\\1")
```



