title: "D Plunkett 607 Week 1"
author: "Diana Plunkett"
date: "1/30/2022"
output: html_document
---
<hr/>  
## Overview

I have surveyed 8 people, asking them to rank 6 movies on a scale of 1-5 (1 = highest and 5 = lowest).

The results are stored in a database called MovieReviews, which has two tables:    
The table Movie has 3 fields:  MovieID, Title, YearReleased    
The table Rating has fields: RatingID, ReviewerID, MovieID, Ranking    

If this had been a survey where there was also some demographic data about the survey respondents, I would have also created a Respondent with the ReviewerID and the demographic info (age, zip code, etc).  But that was not needed here, so best to keep it simple.  

As there were only 6 movies, and that data was static  - if I were to run the survey again, with different people, or the same people in 6 months the same data would apply - I simply loaded the movie data into the Movie table within the script.  This also simplified the replication.  

For the 8 responses, these were saved from a spreadsheet to Movie Survey.csv, and loaded into the database from there.  

Movie Survey.csv is availble on github to be downloaded and used in one's local environment, as is the MySQL script, MovieReviews.    
IMPORTANT: The path for the script must be updated if the Movie Survey.csv is not in c:\data. 

###Load Libraries 
```{r}
library(dbplyr)
library(RMySQL)
library(tidyverse)
library(stringr)
library(magrittr)
```


###Load Database

IMPORTANT NOTE: User should be updated to your username.

```{r}
mydb <- dbConnect(MySQL(), 
  host = "localhost",
  dbname = "MovieReviews",
  user = "newuser",  #REPLACE newuser WITH YOUR USERNAME
  password = rstudioapi::askForPassword("Database password")
)
dbListTables(mydb)
dbGetQuery(mydb, 'SELECT * FROM Movie')
dbGetQuery(mydb, 'select ReviewerID, count(*) from Rating group by ReviewerID')
```

###Load Data to dataframe

This includes a left join to add the name and year of the movie to the ratings.  

```{r}
mv <- dbReadTable(mydb, "Movie")
rt <- dbReadTable(mydb, "Rating")

reviews <- left_join(rt, mv, by = NULL)

print(reviews)

```

###Seeing the results

Also concatenating the Title and YearReleased as I can never remember the order of the Matrix movies, so this helps. 

```{r}

reviews$Title %<>% str_c(" (",  reviews$YearReleased,  ")")

sumReviews <- reviews %>% 
    group_by(Title) %>% 
    summarize(avg_rt = mean(Ranking, na.rm = TRUE), 
              votes = sum(!is.na(Ranking)))

(sumReviews [order(sumReviews$avg_rt),])
```
### Conclusion  

Ready Player One had the highest average rating.  But the Matrix was very very close, and it had 8 votes vs Ready Player One's 6.  

The fewest people voted for (and presumable saw) the The Matrix, Resurections (2021) and those who saw it gave it the lowest ranking.    




